Do NBA Teams Avoid Trading within Their Own Division?

### 2023-02-01 Manuscript.Rmd
Manuscript updates:
Added binary and weighted CMP ERGM results.
Added table with distribution of trade counts.
Added text explaining the above.

Adequate modeling a weighted network via an ERGM requires adapting model specification to the distribution of trade counts across dyads by choosing the reference measure [@krivit2012, sec. 5.2]. While the typical approach involves assuming that the counts are Poisson-distributed, in practice the data is distributed differently with overprevalence of 0s (zero-inflation) or with greater variation than the Poisson distribution assumes (overdispersion). Table \@ref(tab:dyadic-trade-counts) summarises that trading is relatively rare in general and it happens even more rearely that a pair of teams trades more than once in the same season.

@@

At this point one could either choose a non-Poissonian reference measure or simplify data by dychotomizing the counts (any number of trades vs no trades at all) and proceeding with a binary ERGM. To make our conclusions more robust and for illustrative purposes with provide both types of results.

@@

To summarize the results from Figure \@ref(fig:seasonal-ergms) in a single model, we also fit a pooled ERGM across all seasons, including a base-rate effect for each year's trade volume. These results are presented in Table \@ref(tab:model1-table-short).

@@

We find an aggregate same-division homophily effect of 0.0099 (standard error of 0.059, z-statistic of 0.169), again showing no avoidance pattern. Fitting similar models to the periods 1976-2004 and 2004-2019 separately leads to almost identical results.[^periodModels]

@@

@@

@@

```{r model1-table}
withr::with_options(
  list(knitr.kable.NA = ""), {
    knitr::kable(
      readRDS(file = "model1-table.rds"), 
      caption = "Full results for the ERG model fit to pooled data.",
      align = "lrrrr")
  })
```

>>>>

Adequate modeling a weighted network via an ERGM requires adapting model specification to the distribution of trade counts across dyads by choosing the reference measure [@krivit2012, sec. 5.2]. While the typical approach involves assuming that the counts are Poisson-distributed, in practice the data is distributed differently with overprevalence of 0s (zero-inflation) or with greater variation than the Poisson distribution assumes (overdispersion). Table \@ref(tab:dyadic-trade-counts) summarises that trading is relatively rare in general and it happens even more rearely that a pair of teams trades more than once in the same season. Formal tests diagnoze overdispersion in this distribution [@cameron1990regression;@rpkg-AER].

@@

At this point one could either choose a non-Poissonian reference measure or simplify data by dychotomizing the counts (any number of trades vs no trades at all) and proceeding with a binary ERGM. The homophily coefficient in the binary ERGM fit to all seasonal trade networks simultaneously, including season- and division-specific effects, is equal to 0.0122759 (standard error of 0.065, z-statistic of 0.188) showing no avoidance patterns. The complete model is presented in Table \@ref(tab:model-binary-table) in the Appendix.

@@

To summarize the results from Figure \@ref(fig:seasonal-ergms) in a single model, we also fit a pooled ERGM across all seasons, including a base-rate effect for each year's trade volume, similarly to the binary ERGM mentioned above. These results are presented in Table \@ref(tab:model1-table-short).

@@

We find an aggregate same-division homophily effect of 0.0060264 (standard error of 0.059, z-statistic of 0.102), again showing no avoidance pattern. Fitting similar models to the periods 1976-2004 and 2004-2019 separately leads to almost identical results.[^periodModels]

@@

# Model: privaterepo/estimation/ergm-pooled-binary-4
withr::with_options(
  list(knitr.kable.NA = ""), {

@@

  }

@@

```{r model1-table}
withr::with_options(
  list(knitr.kable.NA = ""), {
    structure(list(Effect = c("CMP", "1976-1977", "1977-1978", "1978-1979",  "1979-1980", "1980-1981", "1981-1982", "1982-1983", "1983-1984",  "1984-1985", "1985-1986", "1986-1987", "1987-1988", "1988-1989",  "1989-1990", "1990-1991", "1991-1992", "1992-1993", "1993-1994",  "1994-1995", "1995-1996", "1996-1997", "1997-1998", "1998-1999",  "1999-2000", "2000-2001", "2001-2002", "2002-2003", "2003-2004",  "2004-2005", "2005-2006", "2006-2007", "2007-2008", "2008-2009",  "2009-2010", "2010-2011", "2011-2012", "2012-2013", "2013-2014",  "2014-2015", "2015-2016", "2016-2017", "2017-2018", "2018-2019",  "*Homophily*"), Estimate = c(0.231584299590497, -0.890763342957112,  -0.83536651253216, -0.916362759460457, -0.871853243700118, -0.987841290853209,  -0.900902605442105, -0.800007214754898, -0.899560461864575, -1.0483260921955,  -1.07751539263132, -0.913653500983868, -0.937433706117321, -1.11513540560325,  -1.16287289821496, -1.09836829500166, -1.32117548878016, -1.20705371653326,  -1.28771790721094, -1.39250323356208, -1.22250163107761, -1.22000987502278,  -1.11156405118103, -1.25007055148395, -1.27514745462814, -1.03847047606852,  -1.05882132141435, -1.26269742772427, -1.08833699875934, -1.0469834704777,  -1.07289680849128, -1.1929728670501, -1.1562207254889, -0.961040860349278,  -0.956076839003664, -0.925824485618014, -1.14194414800718, -0.971701230551129,  -0.971414447588866, -0.82982643075161, -1.01829169839617, -1.07774381829707,  -0.999993419528592, -0.912020178092013, 0.00602638250203989),      SE = c(0.12203512128938, 0.0786659348166571, 0.0749602274310118,      0.0789621568846716, 0.0760022797170869, 0.0849915193088094,      0.0765183664206582, 0.069324091908689, 0.0754443838436976,      0.0885417986944042, 0.0913748344030373, 0.0767335950520064,      0.081705154479656, 0.0874539689668034, 0.0838134623779419,      0.0794202098916715, 0.103689830592027, 0.0889138426500533,      0.0980046218251252, 0.105920044602825, 0.0836029172158815,      0.080937063964129, 0.0737605330112484, 0.0872142023820389,      0.0904111796177795, 0.0712261498735629, 0.0719333510361581,      0.0885909751979331, 0.0733621462658981, 0.067206171621419,      0.0674913276175189, 0.0767037480828984, 0.0768718796054209,      0.0627954612107514, 0.06156685684603, 0.0602325458811699,      0.0751034750003797, 0.0636099818174729, 0.0642021297111841,      0.0536773940981808, 0.0669347619251306, 0.0717645838325605,      0.0660829455575422, 0.0579577186366169, 0.0593486673413606     ), "95% CI" = c("(-0.01;  0.47)", "(-1.04; -0.74)", "(-0.98; -0.69)",      "(-1.07; -0.76)", "(-1.02; -0.72)", "(-1.15; -0.82)", "(-1.05; -0.75)",      "(-0.94; -0.66)", "(-1.05; -0.75)", "(-1.22; -0.87)", "(-1.26; -0.90)",      "(-1.06; -0.76)", "(-1.10; -0.78)", "(-1.29; -0.94)", "(-1.33; -1.00)",      "(-1.25; -0.94)", "(-1.52; -1.12)", "(-1.38; -1.03)", "(-1.48; -1.10)",      "(-1.60; -1.18)", "(-1.39; -1.06)", "(-1.38; -1.06)", "(-1.26; -0.97)",      "(-1.42; -1.08)", "(-1.45; -1.10)", "(-1.18; -0.90)", "(-1.20; -0.92)",      "(-1.44; -1.09)", "(-1.23; -0.94)", "(-1.18; -0.92)", "(-1.21; -0.94)",      "(-1.34; -1.04)", "(-1.31; -1.01)", "(-1.08; -0.84)", "(-1.08; -0.84)",      "(-1.04; -0.81)", "(-1.29; -0.99)", "(-1.10; -0.85)", "(-1.10; -0.85)",      "(-0.94; -0.72)", "(-1.15; -0.89)", "(-1.22; -0.94)", "(-1.13; -0.87)",      "(-1.03; -0.80)", "(-0.11;  0.12)"), "p-value" = c(0.058,      0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0,      0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0,      0, 0, 0, 0, 0, 0.919)), row.names = c(NA, -45L), class = "data.frame") |>
    knitr::kable(
      caption = "Full results for the weighted ERG model (Conway-Maxwell-Poisson reference) fit to pooled data.",
      align = "lrrrr")
  })
```
